---
title: "Final consolidation data set analysis"
author: "Kristin Dobbin"
date: "6/23/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
#markdown settings
knitr::opts_knit$set(root.dir = "/Users/KristinDobbin/Dropbox/UCLA/R_Projects_UCLA/Consolidations", warning=FALSE, message=FALSE, include=FALSE)
```

```{r load, include=FALSE, echo= FALSE}
#load libraries
library(tidyverse)

#load data
Data <- read.csv("Outputs/Compiledfinaldata.csv")

#Correct data types
Data$County <- as.factor(Data$County)
Data$System_Type <- as.factor(Data$System_Type) #two NAs are both domestic well communities
Data$Institution_type <- as.factor(Data$Institution_type)
Data$Receiving_System_Institution_Type <- as.factor(Data$Receiving_System_Institution_Type)
Data$SystemID <- as.factor(Data$SystemID)
Data$Receiving_System_ID <- as.factor(Data$Receiving_System_ID)
Data$Issue_summary <- as.factor(Data$Issue_summary)
Data$Issue_capacity <- as.factor(Data$Issue_capacity)
Data$Issue_quality <- as.factor(Data$Issue_quality)
Data$Issue_TMF <- as.factor(Data$Issue_TMF)
Data$Issue_disaster <- as.factor(Data$Issue_disaster)
Data$Issue_sustainability <- as.factor(Data$Issue_sustainability)
Data$Funding_Source <- as.factor(Data$Funding_Source)
Data$Consolidation_Type <- as.factor(Data$Consolidation_Type)
Data$Physical_or_Mangerial <-as.factor(Data$Physical_or_Mangerial)
Data$Mandatory_or_voluntary <- as.factor(Data$Mandatory_or_voluntary)
Data$Consolidatedsystem_primary_source_combinedFINAL <- as.factor(Data$Consolidatedsystem_primary_source_combinedFINAL)
Data$Receiving_2021_source <- as.factor(Data$Receiving_2021_source)
Data$Unique_project <- as.factor(Data$Unique_project) 

#Make subsets for different groups of interest of consolidated systems/types
CWS <- Data %>% filter(!System_Type %in% c('NC', 'NP', 'NTNC'))

Uniqueprojectsa <- Data %>% group_by(Unique_project) %>% filter (! duplicated(Unique_project)) 
Uniqueprojectsa <- Uniqueprojectsa[-1,]
Uniqueprojectsb <- Data %>% filter(Unique_project== "Yes")
Uniqueprojects <- rbind(Uniqueprojectsa, Uniqueprojectsb) #Only use this to summarize project level characteristics like mandatory or physical. Not for Issues (can vary between involved systems and not consolidated system characteristics)

Managerial <- Uniqueprojects %>% filter(Physical_or_Mangerial == "Managerial")
Physical <- Uniqueprojects %>% filter(Physical_or_Mangerial == "Physical")
Mandatory <- Uniqueprojects %>% filter(Mandatory_or_voluntary == "Mandatory")

Receiving_systems <- Data %>% group_by(Receiving_System_ID) %>% filter (! duplicated(Receiving_System_ID)) #Only use this to summarize receiving system information and count of projects by system not project info or consolidated system info
```

## Overview of consolidations 2015-2021

In 2015 there were 7,642 Public Water Systems (PWSs) in the state, 3,015 of which were Community Water Systems (CWS) (2015 Safe Drinking Water Plan). Today, the state counts 7,351 PWS with 2,870 CWS. Some of this reduction can be attributed to system consolidations. Compiling various state data sources we have identified `r nrow(Uniqueprojects)` cases of completed consolidation projects between 2015 and 2016.  

These consolidations have taken place in nearly every county (48 total) and involve the water providers of an estimated `r sum(na.omit(Receiving_systems$Receiving_2021_pop))` Californians. Of them, `r sum(na.omit(Data$Population))` have been incorporated as customers into a new water system as part of a consolidation project, in other words, were previously served by a *consolidated* system. 

Of the `r nrow(Uniqueprojects)` cases we identify, `r nrow(Managerial)` were managerial consolidations, another `r nrow(Physical)` were physical consolidations. Our data set consists of only `r nrow(Mandatory)` mandatory consolidations (Pratt MWC and Ceres West MHP) with the remainder being voluntary consolidations.  

Distance....

## Drivers of consolidation

The SWRCB has publicly available information on factors driving or incentivizing consolidation in `r Data %>% tally(!is.na(Issue_summary))` of the consolidated water systems in our dataset. Grouping this descriptive information into categories, we find that water quality concerns are by far the most common reason for consolidation (`r Data %>% tally(Issue_summary=="Quality")` cases). Table 1 below provides frequencies for all five categories, none indicates that no driving factors or issues were listed. "Sustainability" in turn indicates that the reasons given were pro-active, e.g., reliance on a single source. Capacity indicates source capacity whereas TMF capacity indicates technical, managerial and financial capacity.

```{r drivers, include=TRUE, echo=FALSE}
knitr::kable(summary(na.omit(Data$Issue_summary)), col.names = "#")
```

## Characteristics of all consolidated systems

As preciously mentioned, `r sum(na.omit(Data$Population))` Californians were served by a water system that was consolidated between January 2015 and December 2021. The mean population served by one of these systems is `r round(mean(Data$Population), digits=0)` with a minimum of `r min(Data$Population)` (as listed in SDWIS data) and maximum of `r max(Data$Population)`. The median population served,  however, is `r median(Data$Population)` indicating that the vast majority of consolidated systems are very small. 

The `r nrow(Uniqueprojects)`consolidation projects studied here resulted in the inactivation of approximately `r (nrow(Data)-3)` Public Water Systems between 2015 and 2016. `r Data %>% filter(System_Type == "C") %>% tally()` of these systems were Community Water Systems (CWS). The remainder are a mix of NC and NTNC systems. Further we identify `r Data %>% filter(System_Type == "SSWS") %>% tally()` consolidation of a state small water systems and `r sum(is.na(Data$System_Type))` instances of consolidating domestic well communities. 

`r Data %>% filter(Consolidatedsystem_primary_source_combinedFINAL == "GW" | Consolidatedsystem_primary_source_combinedFINAL == "GWP") %>% tally()` of these systems relied on groundwater as their primary source, another `r Data %>% filter(Consolidatedsystem_primary_source_combinedFINAL == "SW" | Consolidatedsystem_primary_source_combinedFINAL == "SWP") %>% tally()` relied on surface water.

These systems, in turn, represent `r sum(table(unique(Data$Institution_type)))` different institutional/governing types (See Dobbin & Fencl 2021). The most common type is Private (other) signifiying non IOU private systems that could not readily be identified as a mutual water company, homeowners association or mobile home park.The second most common types is Mobile Home Parks. 

```{r gov types, include=TRUE, echo=FALSE}
ConsolidatedtypeALL <- as.data.frame(table(Data$Institution_type))
ConsolidatedtypeALL <- rename(ConsolidatedtypeALL, "Institutional_Type" = "Var1")
ConsolidatedtypeALL <- ConsolidatedtypeALL[order(ConsolidatedtypeALL$Freq, decreasing = TRUE),] 
ConsolidatedtypeALL$Freq[ConsolidatedtypeALL$Freq==0] <- NA
ConsolidatedtypeALL<-ConsolidatedtypeALL[complete.cases(ConsolidatedtypeALL),]

knitr::kable(ConsolidatedtypeALL, row.names=FALSE)
```


## Characteristics of consolidated Community Water Systems

Among just the consolidated systems that were CWSs, these statistics are as follows. 

```{r CWS consolidated systems, include=TRUE, echo=FALSE}
knitr::kable(summarize(CWS,
           Count = n(),
           Population_mean = mean(Population),
           Population_median = median(Population),
           Population_min = min(Population),
           Population_max = max(Population)))
```

`r CWS %>% filter(Consolidatedsystem_primary_source_combinedFINAL == "GW" | Consolidatedsystem_primary_source_combinedFINAL == "GWP") %>% tally()` of the consolidated CWS relied on groundwater as their primary source, another `r CWS %>% filter(Consolidatedsystem_primary_source_combinedFINAL == "SW" | Consolidatedsystem_primary_source_combinedFINAL == "SWP") %>% tally()` relied on surface water.   

These systems, in turn, represent `r sum(table(unique(CWS$Institution_type)))` different institutional/governing types (See Dobbin & Fencl 2021). The most common type is Mobile Home Parks followed by IOUs and "Private (other)" systems, representing non IOU private systems that could not readily be identified as a mutual water company, homeowners association or mobile home park.

```{r gov types, include=TRUE, echo=FALSE}
ConsolidatedtypeCWS <- as.data.frame(table(CWS$Institution_type))
ConsolidatedtypeCWS <- rename(ConsolidatedtypeCWS, "Institutional_Type" = "Var1")
ConsolidatedtypeCWS <- ConsolidatedtypeCWS[order(ConsolidatedtypeCWS$Freq, decreasing = TRUE),] 
ConsolidatedtypeCWS$Freq[ConsolidatedtypeCWS$Freq==0] <- NA
ConsolidatedtypeCWS<-ConsolidatedtypeCWS[complete.cases(ConsolidatedtypeCWS),]

knitr::kable(ConsolidatedtypeCWS, row.names=FALSE)
```

Demographics.....

## Characteristics of receiving water systems
Among the `r nrow(Receiving_systems)` unique receiving water systems, 87 rely on groundwater as their primary source, another 55 rely on surface water. 

Type...

```{r, receiving system characteristics, include=TRUE, echo=FALSE}
Projectbyreceiving <- as.data.frame(table(Uniqueprojects$Receiving_System_ID))
```

`r Projectbyreceiving %>% filter(Freq >1) %>% tally()` of these systems participated in more than one consolidation project over the study period. `r Projectbyreceiving %>% filter(Freq == 6) %>% tally()` systems participated in 6 consolidation projects each over the course of the 6 years. Another system, South Tahoe PUD, participated in `r max(Projectbyreceiving$Freq)`.

## Characteristics of only systems "receiving" CWSs in consolidations

Among the systems that were receiving systems in the above cases of CWS consolidation, we find similar diversity in institutional types, IOUs are the most common receiving system type and City's the second most common. Overall, `r sum(table(unique(CWS$Receiving_System_Institution_Type)))` different institutional types acted as receiving systems over the studied period. 

```{r recieving systems, include=TRUE, echo=FALSE}
Receivingtype <- as.data.frame(table(CWS$Receiving_System_Institution_Type))
Receivingtype <- rename(Receivingtype, "Institutional_Type" = "Var1")
Receivingtype <- Receivingtype[order(Receivingtype$Freq, decreasing = TRUE),] 
Receivingtype$Freq[Receivingtype$Freq==0] <- NA
Receivingtype<-Receivingtype[complete.cases(Receivingtype),]

knitr::kable(Receivingtype, row.names=FALSE)
```

## Governance change as a result of CWS consolidations

```{r gov combos, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
Pathways <- CWS
Pathways$Pathway <- paste(Pathways$Institution_type, ", ", Pathways$Receiving_System_Institution_Type)
Pathways$Pathway <- as.factor(Pathways$Pathway)
Pathwaystable <- as.data.frame(table(Pathways$Pathway))
Pathwaystable <- Pathwaystable[order(Pathwaystable$Freq, decreasing = TRUE),]
```

Using this information about consolidated and receiving system types, then, we are able to identify how CWS governance is changing in the state as a result of consolidations. In total, CWS consolidations represent `r sum(table(unique(Pathwaystable$Var1)))` distinct governance changes, the five most common of which are displayed in the table below. 

```{r, most common combos, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
Pathwaystable_head <- top_n(Pathwaystable, 5)
knitr::kable(Pathwaystable_head, row.names=FALSE, col.names = c("Change from, to", "#"))
```

## Demographic comparison between consolidated CWS and receiving systems

demographics.....

## Appendix

### Institutional types for all consolidated water systems (not just CWS):

```{r appendix1, include=TRUE, echo=FALSE}
Consolidatedtype <- as.data.frame(table(Data$Institution_type))
Consolidatedtype <- rename(Consolidatedtype, "Consolidated_Inst_Type" = "Var1")
Consolidatedtype <- Consolidatedtype[order(Consolidatedtype$Freq, decreasing = TRUE),] 

knitr::kable(Consolidatedtype, row.names=FALSE)
```

### Institutional types for all receiving water systems (not just those receiving CWS systems):

```{r appendix2, include=TRUE, echo=FALSE}
Receivingtype_all <- as.data.frame(table(Data$Receiving_System_Institution_Type))
Receivingtype_all <- rename(Receivingtype_all, "Institutional_Type" = "Var1")
Receivingtype_all <- Receivingtype_all[order(Receivingtype_all$Freq, decreasing = TRUE),] 

knitr::kable(Receivingtype_all, row.names=FALSE)
```

### Top ten governance change pathways for all consolidations cases (not just CWS)

```{r appendix3, include=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
Pathways_all <- Data
Pathways_all$Pathway <- paste(Pathways_all$Institution_type, ", ", Pathways_all$Receiving_System_Institution_Type)
Pathways_all$Pathway <- as.factor(Pathways_all$Pathway)
Pathwaystable_all <- as.data.frame(table(Pathways_all$Pathway))
Pathwaystable_all <- Pathwaystable_all[order(Pathwaystable_all$Freq, decreasing = TRUE),]
Pathwaystable_all_head <- top_n(Pathwaystable_all, 10)
knitr::kable(Pathwaystable_all_head, row.names=FALSE, col.names = c("Change from, to", "#"))
```

